---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%"
)
set.seed(123456)
```

# `fwb`: Fractional Weighted Bootstrap

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/fwb)](https://CRAN.R-project.org/package=fwb)
<!-- badges: end -->

`fwb` implements the fraction weighted bootstrap, also known as the Bayesian bootstrap, following the treatment by Xu et al. (2020). The primary function is `fwb()`, which is essentially a drop-in for `boot::boot()` in that it takes in a dataset and a function and applies that function to the dataset, generating sets of weights from a uniform Dirichlet distribution to be applied to the statistic(s) computed in the function. Also included are `fwb.ci()`, a drop-in for `boot::boot.ci()` for computing various kinds of confidence intervals (e.g., percentile, normal, bias-corrected percentile, etc.), and `vcovFWB()`, a drop-in for `sandwich::vcovBS()` for computing a parameter covariance matrix from a regression model using the fractional weighted bootstrap.

Check out the `fwb` [website](https://ngreifer.github.io/fwb)!

## Installation

You can install the development version of `fwb` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ngreifer/fwb")
```

## Examples

Below are some examples of how to use `fwb`.

```{r}
library("fwb")
```

### Bearing cage field failure Weibull analysis from Xu et al. (2000)

This example involves performing a Weibull analysis to estimate the $\beta$ (shape) parameter of the Weibull distribution characterizing the time to failure of a set of aircraft engines. Among the 1703 engines, there were only 6 failures and 1697 right-censored observations. The traditional (resampling-based) bootstrap would fail if a bootstrap replication omitted the 6 failures, but all failures are retained when using the fractional weighted bootstrap, which makes it particularly effective for this analysis.

```{r, fig.align='center', fig.height=3, fig.width=7}
data("bearingcage", package = "fwb")

# Function to compute the scale (eta) and shape (beta) parameters
# from weighted data
weibull_est <- function(data, w) {
  fit <- survival::survreg(survival::Surv(hours, failure) ~ 1,
                           data = data, weights = w,
                           dist = "weibull")

  c(eta = unname(exp(coef(fit))), beta = 1/fit$scale)
}

# 1999 bootstrap replications; more is always better
fwb_est <- fwb(bearingcage, statistic = weibull_est,
                R = 1999, verbose = FALSE)
fwb_est

# Bias-corrected percentile confidence interval
summary(fwb_est, ci.type = "bc")

# Plot the bootstrap distribution
plot(fwb_est, index = "beta", qdist = "chisq")
```

### Infertility logistic regression analysis using `infert` dataset

This example demonstrates using `vcovFWB()` to estimate standard errors for the effect of spontaneous and induced abortions on infertility as analyzed in Trichopoulos et al. (1976). Patients are organized into matched sets of 3 patients each. We use a fixed effects logistic regression to adjust for matched set membership (ignoring the potential bias in this approach for the sake of the example). The traditional bootstrap fails because many matched sets will fully omit either cases or non-cases, leading to perfect prediction and the failure of the model to converge, yielding invalid estimates. Because all units are retained when using the fractional weighted bootstrap, the model always converges and the estimates are reasonable.

```{r, message=F, warning=F}
data("infert")

fit <- glm(case ~ spontaneous + induced + factor(stratum),
           data = infert, family = quasibinomial())

library("lmtest")

# The traditional bootstrap fails
coeftest(fit, vcov = sandwich::vcovBS)[1:3,]

# The fractional weighted bootstrap succeeds
coeftest(fit, vcov = vcovFWB)[1:3,]
```

We can also perform cluster-robust inference by bootstrapping the strata. (Note in this case the traditional bootstrap does fine, but the fractional weighted bootstrap is still more accurate.)

```{r}
# Including stratum membership as a clustering variable
coeftest(fit, vcov = vcovFWB, cluster = ~stratum)[1:3,]
```

Let's look more in-depth at the results of the traditional and fractional weighted bootstrap by comparing the output of `fwb()` and `boot::boot()`.

```{r, warning=F}
fit_fun <- function(data, w) {
  fit <- glm(case ~ spontaneous + induced + factor(stratum),
           data = data, weights = w, family = quasibinomial())
  coef(fit)[1:3]
}

boot_est <- boot::boot(infert, fit_fun, R = 999, stype = "f")
boot_est

fwb_est <- fwb(infert, fit_fun, R = 999, verbose = FALSE)
fwb_est
```

Already the bias and standard errors indicate problems with the traditional bootstrap. Let's plot histograms of the estimates to see where the failure is:

```{r, fig.align='center', fig.height=3, fig.width=7}
plot(boot_est, index = 2)
```

```{r, fig.align='center', fig.height=3, fig.width=7}
plot(fwb_est, index = 2)
```

It is clear that the estimates from the traditional bootstrap are pathological, whereas the estimates from the fractional weighted bootstrap are more reasonable. The non-normality of the fractional weighted bootstrap distributions also suggests that the usual Wald-style confidence intervals may not be accurate, and a percentil interval should be computed instead.

## When to use the fractional weighted bootstrap

The fractional weighted bootstrap is uniformly more reliable than the traditional bootstrap when a weighted statistic can be computed (though this doesn't mean the bootstrap is always valid). In most simple case, both methods will yield the same results. In some pathological examples like those above, the fractional weighted bootstrap dramatically outperforms the traditional bootstrap. This will be true when running regression models with sparse categorical variables either in the outcome or among the predictors, for example, when estimating fixed effects or when a binary outcome is rare. However, it is important to know when a weighted statistic can be computed; for example, computing the weighted median is not always straightforward, making the traditional bootstrap potentially more useful for computing it. Still, though, the fractional weighted bootstrap deserves a place in an analysts toolbox.

## Related packages

* `boot`, which provides the traditional bootstrap, including an interface that accepted frequency weights to compute weighted statistics, as was used above
* `bayesboot`, which also provides functionality for the Bayesian bootstrap but does so in a more explicitly Bayesian fashion and with returned objects that are less consistent with those from `boot`

## Author

* Noah Greifer (noah.greifer\@gmail.com)

## References

Trichopoulos, D., Handanos, N., Danezis, J., Kalandidi, A. and Kalapothaki, V. (1976), Induced Abortion and Secondary Infertility. *BJOG: An International Journal of Obstetrics & Gynaecology*, 83, 645-650. https://doi.org/10.1111/j.1471-0528.1976.tb00904.x

Xu, L., Gotwalt, C., Hong, Y., King, C. B., & Meeker, W. Q. (2020). Applications of the Fractional-Random-Weight Bootstrap. *The American Statistician*, 74(4), 345–358. https://doi.org/10.1080/00031305.2020.1731599
